---
title: "Week 2 Exercises"
author: "Auriana Anderson"
date: "October 28, 2024"
output:
 pdf_document:
   latex_engine: lualatex
---

Please complete all exercises below. You may use stringr, lubridate, or the forcats library. 

Place this at the top of your script:
library(stringr)
library(lubridate)
library(forcats)

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

options(tinytex.verbose = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/aanderson/Desktop/Intro to R and Python/DES5002 Intro to R and Python")

```
# Exercise 1
Read the sales_pipe.txt file into an R data frame as sales. 

```{r Read file in}
library(stringr)
library(lubridate)
library(forcats)

# setwd("C:/Users/aanderson/Desktop/Intro to R and Python/DES5002 Intro to R and Python")

sales <- read.delim("Week_2/Data/sales_pipe.txt",
                    stringsAsFactors=FALSE,
                     sep = "|", fileEncoding = "latin1")

```

# Exercise 2
You can extract a vector of columns names from a data frame using the colnames() function. Notice the first column has some odd characters. Change the column name for the FIRST column in the sales date frame to Row.ID. 

**Note: You will need to assign the first element of colnames to a single character.**

```{r change first column}

colnames(sales)

names(sales)[names(sales) == "ï..Row.ID"] <- "Row.ID" 

head(sales)
```

# Exercise 3
Convert both Ship.Date and Order.Date to date vectors within the sales data frame. What is the number of days between the most recent order and the oldest order?
How many years is that?
How many weeks?

**Note: Use lubridate**

```{r ship and order to vector}

sales$Ship.Date <- mdy(sales$Ship.Date)

sales$Order.Date <- mdy(sales$Order.Date)

oldest_order <- min(sales$Order.Date)

recent_order <-max(sales$Order.Date)

#What is the number of days between the most recent order and the oldest order?
days_between <- as.numeric(recent_order-oldest_order)

#How many years is that?
years_between <- days_between/365.25

#How many weeks?
weeks_between <- days_between/7

oldest_order
recent_order
days_between
years_between
weeks_between

```


# Exercise 4
What is the average number of days it takes to ship an order?

```{r num days to ship}

shiptime <-as.numeric(sales$Ship.Date - sales$Order.Date)

avg_shiptime <- mean(shiptime)
avg_shiptime

```
# Exercise 5
How many customers have the first name Bill?
You will need to split the customer name into first and last name segments and then use a regular expression to match the first name bill. Use the length() function to determine the number of customers with the first name Bill in the sales data. 

```{r Bill}

name_split <- str_split(string = sales$Customer.Name, pattern = " ")

sales$first_name <- sapply(name_split, "[", 1)

number_of_bills <- sum(length(which(sales$first_name == "Bill")))
number_of_bills

```

# Exercise 6
How many mentions of the word 'table' are there in the Product.Name column? 

There are zero occurrences of table.

There are 230 occurrences of Table.

capitalization seems to make a difference here.

**Note you can do this in one line of code**
```{r count table}

#I looked this up and found in order to find just the word table without 
#it being apart of a longer word and having exact matches you must use the 
#regex \\b on either side of the word

sum(str_count(sales$Product.Name, "\\bTable\\b"))

```

# Exercise 7
Create a table of counts for each state in the sales data. The counts table should be ordered alphabetically from A to Z. 

```{r table of counts}

count_state <-as.data.frame(table(sales$State))

colnames(count_state) <- c("State","Count")

count_state <- count_state[order(count_state$State),]

head(count_state)

```

# Exercise 8
Create an alphabetically ordered barplot for each sales Category in the State of Texas. 

```{r texas only}

texas <- sales[sales$State == "Texas",]

head(texas)

texas_count <- table(texas$Category)

texas_count <- as.data.frame(texas_count)
```


```{R texas names}
colnames(texas_count) <- c("Category","Count")

texas_count <- texas_count[order(names(texas_count))]

texas_count
```


```{R alphabarplot}

barplot(texas_count$Count,
        names.arg = texas_count$Category,
        col = "pink",
        main = "Sales Category Counts in Texas"
        ,font.main = 4)

```

# Exercise 9
Find the average profit by region.
**Note: You will need to use the aggregate() function to do this. To understand how the function works type ?aggregate in the console.**

```{r avg prof region}

aggregate(x = list(Avg_Profit = sales$Profit), by = list(Region = sales$Region), FUN = "mean")

```

# Exercise 10
Find the average profit by order year. 
**Note: You will need to use the aggregate() function to do this. To understand how the function works type ?aggregate in the console.**

```{r avg prof orderyr}

date_split <-  str_split(string = sales$Order.Date, pattern = "-")

sales$order_year <- sapply(date_split, "[", 1)

aggregate(x = list(Avg_Profit = sales$Profit), by = list(order_year = sales$order_year), FUN = "mean")

```

